---
title: "Global Historical Temperatures"
author: "Rashan Jibowu"
date: "08/02/2015"
output:
  html_document:
    keep_md: yes
---

## Introduction

x

## Set Up Environment

```{r}

library(plyr) 
library(ggplot2)

```

## Load the data

```{r load}

# Data includes NA encodings as '***' and '****'
data <- read.csv("./data/ExcelFormattedGISTEMPDataCSV.csv", na.strings = c('***','****'))

```

Let's checkout the structure of the data

```{r structure}

str(data)

```

## Clean the data

Considering the struture of the data observed above, we'll need to do 2 things to make this data more usable:

1. Fix the formatting issue where several months are encoded as factors rather than integers
2. Clarify what `J.D`, `D.N`, `DJF`, `MAM`, `JJA`, and `SON` mean

### Fix Formatting

```{r clean}

# fix formatting (only useful when you don't include the na.strings argument when reading in the data)
data$Jul <- as.integer(data$Jul)
data$Aug <- as.integer(data$Aug)
data$Sep <- as.integer(data$Sep)
data$Oct <- as.integer(data$Oct)
data$Nov <- as.integer(data$Nov)
data$Dec <- as.integer(data$Dec)
data[['J.D']] <- as.integer(data[['J.D']])
data[['D.N']] <- as.integer(data[['D.N']])
data[['DJF']] <- as.integer(data[['DJF']])
data[['JJA']] <- as.integer(data[['JJA']])
data[['SON']] <- as.integer(data[['SON']])

# check the structure again
str(data)

```

### Rename variables

According to the data source, NASA's Goddard Institute for Space Studies, its [FAQs](http://data.giss.nasa.gov/gistemp/FAQ.html) explain what the shorthand variables mean.

> When comparing seasonal temperatures, it is convenient to use "meteorological seasons" based on temperature and defined as groupings of whole months. Thus, Dec-Jan-Feb is the Northern Hemisphere meteorological winter, Mar-Apr-May is N.H. meteorological spring, Jun-Jul-Aug is N.H. meteorological summer and Sep-Oct-Nov is N.H. meteorological autumn. String these four seasons together and you have the meteorological year that begins on Dec. 1 and ends on Nov. 30.

So, let's encode accordingly

```{r rename}

data <- rename(data, c('DJF'='Winter', 
                       'MAM'='Spring', 
                       'JJA'='Summer', 
                       'SON'='Autumn',
                       'J.D'='CalYear',
                       'D.N'='MetYear'))

str(data)

```

## Explore the data

#### Deviations in Calendar Year

```{r explore}

ggplot(data, aes(x = Year, y = CalYear)) + geom_line()

```

#### Deviations in Meteorological Year

```{r explore2}

ggplot(data, aes(x = Year, y = MetYear)) + geom_line()

```

#### Deviations in Meteorological Year (Scatter Plot)

```{r explore3}

ggplot(data, aes(x = Year, y = MetYear)) + geom_point()

```


#### Deviations by Season

```{r explore4}

ggplot(data, aes(x = Year)) + 
    geom_line(aes(y = Winter, colour = "Winter")) + 
    geom_line(aes(y = Summer, colour = "Summer")) +
    geom_line(aes(y = Autumn, colour = "Autumn")) +
    geom_line(aes(y = Spring, colour = "Spring"))

```

#### Deviations by Season (Scatter)

```{r explore5}

ggplot(data, aes(x = Year)) + 
    geom_point(aes(y = Winter, colour = "Winter")) + 
    geom_point(aes(y = Summer, colour = "Summer")) +
    geom_point(aes(y = Autumn, colour = "Autumn")) +
    geom_point(aes(y = Spring, colour = "Spring"))

```

## Visualize the data

Specify a year range and a "season" to display

```{r visualize}

visualize <- function(yearRange, season = "CalYear") {
    
   ggplot(data[data$Year %in% yearRange, ], aes(x = Year)) +
       aes_string(y = season) + 
       geom_line()
}

visualize(seq(1880, 2010, 5), "MetYear")

```

## Write the data for browser based visualization

```{r write}

write.csv(data, "./data/plotdata.csv")

```